Sampling requirements to detect ecosystem change

### 2020-02-26 R/manuscript_draft.Rmd
add some text

\vspace{5 mm}
Keywords: {time-series} \vspace{10 mm}

@@

Because of new technological advances, there are many data sources that can be derived long after the actual  processes occurred. For example, paleoecological lake cores can be extracted and historical ecological communities can be reconstructured (citations from Rosalie). Similarly, environmental samples (e.g. water, soil) can be saved and processed later for composition, including eDNA (citation). Similarly, video-graphical approaches can record snapshots of a system and be analyzed later (citations). In each of these cases, decisions have to made about how much data to extract from the previously collected samples. Should the paleoecological core be analyzed at every centimeter? Should the video be assessed at every minute? As long as processing samples is expensive, these trade-offs will remain.

Here, we develop a set of tools to determine the appropriate number of samples and sampling approach when dealing with data sources where additional samples can be added. We focus on paleoecological core samples as one example of this type of data. We examine the situation where the goal is to detect the time at which a significant change in an ecological community occurs, a changepoint. However, our approaches are widely applicable to other questions and data types. We first investigate these tools using a simulated-based approach. As a case study, we apply these tools to the detection of change points for a set of lakes. (more from Rosalie here).

# Sampling approaches and changepoint detection

@@

Previously, Cladoceran assemblage was reconstructed at a 2-3 years resolution, from a 74-cm sediment core (VAR10-10) covering the 1816(±26)—2010 period [@bruel_seeking_2018]. We detected changepoints on the first component computed from the assemblage data. We used the same ordination method as in Bruel et al (2018), but note that Detrended Correspondence Analysis (DCA) is best used to conserve multiple gradients (\easton{Rosalie: can you explain this a bit more, I'm not sure what it means}). As a result, several changepoints were detected on the first component (in 1926/1928, 1946/1948 and 1983/1988). We defined these as the "true" changepoints given they came from an analysis of the complete data. The 1983-1988 changepoint was the largest in magnitude. (\easton{Rosalie: does this mean that the changepoint occurred sometime between 1983 and 1988} \rosalie{The changepoint was detected in 1988, but in fact, it could have taken place anytime between 1983 and 1988. That is just a way to see it, but let me know if it is confusing. I reviewed a paper and asked the authors to change the changepoint to this bracket, and they replied telling me they didn't understand my comment, so maybe it is not worth the confusion! Put another way, if we were recording the temperature every hour, and it was 6°C at 1, 2, 3, 4 am, then 8°C at 5, 6, 7, 8am, the method would detect the changepoint at 5am. But in fact, the temperature could have very well changed anytime between 4.01am and 5am. So I would say: "temperature changed between 4 and 5 am.". Am I maling any sense?}). <!--We conducted the analysis on the second component as well, where only one changepoint was detected (in 1944/1946).-->

@@

>>>>

\vspace{5 mm}
Keywords: {time-series} {changepoint} \vspace{10 mm}

@@

Because of new technological advances, there are many data sources that can be derived long after the actual  processes occurred. For example, paleoecological lake cores can be extracted and historical ecological communities can be reconstructured (citations from Rosalie). Similarly, environmental samples (e.g. water, soil) can be saved and processed later for composition, including eDNA (citation). Likewise, video-graphical approaches can record snapshots of a system and be analyzed later (citations). In each of these cases, decisions have to made about how much data to extract from the previously collected samples. Should the paleoecological core be analyzed at every centimeter? Should the video be assessed at every minute? As long as processing samples is expensive, these trade-offs will remain.

Here, we develop a set of tools to determine the appropriate number of samples and sampling approach when dealing with data sources where additional samples can be added. We tailor our analysis to the detection of changepoint. Changepoint are an important characteristic of a time serie We focus on paleoecological core samples as one example of this type of data. We examine the situation where the goal is to detect the time at which a significant change in an ecological community occurs, a changepoint. However, our approaches are widely applicable to other questions and data types. We first investigate these tools using a simulated-based approach. We then test the tools on a case study from a paleosequence. 
# Sampling approaches and changepoint detection

@@

Previously, Cladoceran assemblage was reconstructed at a 2-3 years resolution, from a 74-cm sediment core (VAR10-10) covering the 1816(±26)—2010 period [@bruel_seeking_2018]. We detected changepoints on the first component computed from the assemblage data. We used the same ordination method as in Bruel et al (2018), but note that Detrended Correspondence Analysis (DCA) is best used to conserve multiple gradients (\easton{Rosalie: can you explain this a bit more, I'm not sure what it means}). As a result, several changepoints were detected on the first component (in 1926/1928, 1946/1948 and 1983/1988). We defined these as the "true" changepoints given they came from an analysis of the complete data. The 1983-1988 changepoint was the largest in magnitude. (\easton{Rosalie: does this mean that the changepoint occurred sometime between 1983 and 1988.} \rosalie{Answer: The changepoint was detected in 1988, but in fact, it could have taken place anytime between 1983 and 1988. That is just a way to see it, but let me know if it is confusing. I reviewed a paper and asked the authors to change the changepoint to this bracket, and they replied telling me they didn't understand my comment, so maybe it is not worth the confusion! Put another way, if we were recording the temperature every hour, and it was 6°C at 1, 2, 3, 4 am, then 8°C at 5, 6, 7, 8am, the method would detect the changepoint at 5am. But in fact, the temperature could have very well changed anytime between 4.01am and 5am. So I would say: "temperature changed between 4 and 5 am.". Am I maling any sense?}). <!--We conducted the analysis on the second component as well, where only one changepoint was detected (in 1944/1946).-->

@@

Many ecological processes are non linear (e.g., D'Amario et al 2019, Scientific reports). Detecting a changepoint on one aspect of the ecosystem does not guarantee that the sampling resolution is appropriate to detect the driver for the changepoint.
